---
title: "Convert Phylogeny"
output: html_document
editor_options: 
  chunk_output_type: console
---

Once the phylogenies have been generated (by `01_generate-phylogeny.Rmd`),
we have then to convert them 
according to the different representations considered.
In our study we considered four different phylogeny representations:

1.  summary statistics
2.  CBLV (matrix encoding of the whole phylogeny)
3.  LTT (Lineage Through Time)
4.  graph

We go over the conversion toward each of these representations one by one.

## Set up 

```{r}
source("R/phylo-inference-ml.R")
set.seed(113)
phylo <- readRDS("data/new-phylogeny-crbd.rds") # change file name if needed
true_params <- readRDS("data/new-phylogeny-crbd-true.rds") # same
```

## Summary statistics

The computation of the summary statistics can be done simply by calling 
`generateSumStatFromPhylo` with the phylogenies and the true parameters given
as arguments. 

```{r}
sumstat <- generateSumStatFromPhylo(phylo, true_params) 
```

Saving

```{r}
saveRDS(sumstat, "data/new-phylogeny-crbd-sumstat.rds")
```

## CBLV

To encode the phylogeny call `generate_encoding`.
Moreover, if you are considering the BiSSE model
and you want to include tip state information in the encoding,
call instead `generate_encoding_bisse`.

```{r}
max_taxa <- 1000 # maximum phylogeny size, change if needed
cblv <- generate_encoding(phylo, max_taxa) 
# If model == "bisse", do call 
# cblv <- generate_encoding_bisse(phylo, max_taxa)
```

Saving 

```{r}
saveRDS(sumstat, "data/new-phylogeny-crbd-cblv.rds")
```


## LTT

LTT or Lineage Through Time is a timeserie 
describing the increase of phylogeny lineages over time.
In a similar manner than above representations,
LTTs can be generated in one line.

```{r}
taxa_range <- c(100,1000) # range of phylogeny size
ltt <- generate_ltt_dataframe(phylo, taxa_range, true_params)$ltt
```

Saving 

```{r}
saveRDS(sumstat, "data/new-phylogeny-crbd-ltt.rds")
```


## Graph

The last representation is the phylogeny graph.
As phylogeny graphs are handled by PyTorch Geometric (in Python), 
this step is just a pre-formatting of the data. 
The formatting will be finalized in Python.

```{r}
graphs <- generate_phylogeny_graph(phylo)
```

Saving 

```{r}
saveRDS(graphs, "data/new-phylogeny-crbd-graph.rds")
```

