---
title: "Predict phylogeny parameters with DNN and Summary statistics"
output: html_document
editor_options: 
  chunk_output_type: console
---

Before running this you should have 
generated your phylogenies (`01_generate-phylogeny.Rmd`)
and computed their summary statistics (`02_convert-phylogeny.Rmd`).

## Set up

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../")
```

Data preparation: here we just need to scale the summary statistics,
before giving them to the DNN,
so that they are all put on an equal footing. 

```{r}
source("R/phylo-inference-ml.R")
set.seed(113)
sumstat <- readRDS("data/new-phylogeny-crbd-sumstat.rds")
true <- readRDS("data/new-phylogeny-crbd-true.rds")
n_taxa <- c(100, 1000) # range of phylogeny size
sumstat <- scale_summary_statistics(sumstat, c(100,1000), names(true))
device = "cpu" # change if you want to compute GPUs
```


## Create datasets

Define the of the training, validation and test sets. 

```{r}
# Define size of datasets.
n_train    <- 90
n_valid    <- 5
n_test     <- 5
batch_size <- 4

# Pick the phylogenies randomly.
ds <- convert_ss_dataframe_to_dataset(sumstat)
train_indices <- sample(1:nrow(sumstat), n_train)
not_train_indices <- setdiff(1:nrow(sumstat), train_indices)
valid_indices <- sample(not_train_indices, n_valid)
test_indices  <- setdiff(not_train_indices, valid_indices)

# Create the datasets.
train_ds <- ds(sumstat[train_indices, ], names(true), c())
valid_ds <- ds(sumstat[valid_indices, ], names(true), c())
test_ds  <- ds(sumstat[test_indices, ], names(true), c())

# Create the dataloader.
train_dl <- train_ds %>% dataloader(batch_size=batch_size, shuffle=TRUE)
valid_dl <- valid_ds %>% dataloader(batch_size=batch_size, shuffle=FALSE)
test_dl  <- test_ds  %>% dataloader(batch_size=1, shuffle=FALSE)
```

## Build the neural network 

```{r}
# Specify neural network parameters. 
n_in      <- length(train_ds[1]$x) # number of neurons of the input layer 
n_out     <- length(true)
n_hidden  <- 100 # number of neurons in the hidden layers 
p_dropout <- 0.01 # dropout probability 
n_epochs  <- 100 # maximum number of epochs for the training 
patience  <- 4 # patience of the early stopping 

# Build the neural network.
dnn.net <- nn_module(
  
  "ss-dnn", 
  
  initialize = function(){
    self$fc1 <- nn_linear(in_features = n_in, out_features = n_hidden)
    self$fc2 <- nn_linear(in_features = n_hidden, out_features = n_hidden)
    self$fc3 <- nn_linear(in_features = n_hidden, out_features = n_hidden)
    self$fc4 <- nn_linear(in_features = n_hidden, out_features = n_hidden)
    self$fc5 <- nn_linear(in_features = n_hidden, out_features = n_out)
  }, 
  
  forward = function(x){
    x %>%
      self$fc1() %>%
      nnf_relu() %>%
      nnf_dropout(p = p_dropout) %>%
      
      self$fc2() %>%
      nnf_relu() %>%
      nnf_dropout(p = p_dropout) %>%
      
      self$fc3() %>%
      nnf_relu() %>%
      nnf_dropout(p = p_dropout) %>%
      
      self$fc4() %>%
      nnf_relu() %>%
      nnf_dropout(p = p_dropout) %>%
      
      self$fc5()
  }
)

# Set up the neural network.
dnn <- dnn.net() # create CNN
dnn$to(device = device) # Move it to the choosen GPU
opt <- optim_adam(params = dnn$parameters) # optimizer 
```

## Training 

```{r}
train_batch <- function(b){
  opt$zero_grad()
  output <- dnn(b$x$to(device = device))
  target <- b$y$to(device = device)
  loss <- nnf_mse_loss(output, target)
  loss$backward()
  opt$step()
  loss$item()
}

valid_batch <- function(b) {
  output <- dnn(b$x$to(device = device))
  target <- b$y$to(device = device)
  loss <- nnf_mse_loss(output, target)
  loss$item()
}
```

```{r}
# Initialize parameters for the training loop.
epoch     <- 1
trigger   <- 0 
last_loss <- 100


# Training loop.
while (epoch < n_epochs & trigger < patience) {
  
  # Training 
  dnn$train()
  train_loss <- c()
  coro::loop(for (b in train_dl) { # loop over batches 
    loss <- train_batch(b)
    train_loss <- c(train_loss, loss)
  })
  
  # Print Epoch and value of Loss function 
  cat(sprintf("epoch %0.3d/%0.3d - train - loss: %3.5f \n",
              epoch, n_epochs, mean(train_loss)))
  
  # Validation 
  dnn$eval()
  valid_loss <- c()
  coro::loop(for (b in test_dl) { # loop over batches 
    loss <- valid_batch(b)
    valid_loss <- c(valid_loss, loss)
  })
  current_loss <- mean(valid_loss)
  
  # Early Stopping 
  if (current_loss > last_loss){trigger <- trigger + 1} 
  else{
    trigger   <- 0
    last_loss <- current_loss
  }
  
  # Print Epoch and value of Loss function
  cat(sprintf("epoch %0.3d/%0.3d - valid - loss: %3.5f \n",
              epoch, n_epochs, current_loss))
  
  epoch <- epoch + 1 
}
```

## Evaluation 

Compute predicted parameters on test set.

```{r}
dnn$eval()
pred <- vector(mode = "list", length = n_out)
names(pred) <- names(true)

# Compute predictions 
coro::loop(for (b in test_dl) {
  out <- dnn(b$x$to(device = device))
  p <- as.numeric(out$to(device = "cpu")) # move the tensor to CPU 
  for (i in 1:n_out){pred[[i]] <- c(pred[[i]], p[i])}
})
```

Now that you have the predicted parameters you can, for instance, 
plot the predicted value by the neural network vs. the true values.

```{r}
par(mfrow=c(1,2))
plot(true[[1]][test_indices], pred[[1]])
abline(0, 1)
plot(true[[2]][test_indices], pred[[2]])
abline(0, 1)
```

